---
title: "Project 1"
author: "Nymph Kong"
format: html
editor: visual
---

```{r}
#| include: false
library(tidytuesdayR)
library(tidyverse)
library(dplyr)
library(here)
library(readxl)
library(janitor)

tuesdata <- tt_load('2025-06-24')

measles <- tuesdata$measles
cases_month <- tuesdata$cases_month
cases_year <- tuesdata$cases_year
```

## Data Context

This data contains measles and rubella cases across the world. This data was downloaded from the **World Health Organisation [Provisional monthly measles and rubella data](https://immunizationdata.who.int/global?topic=Provisional-measles-and-rubella-data&location=) on 2025-06-12**. Measles is a worldwide threat across many countries, with varying degrees of outbreaks. The data collects measles and rubella cases: clinically-compatible, epidemiologically linked, and laboratory-confirmed across 194 countries and separates the data into two data frames, monthly and yearly statistics.

Here are the first six rows of the monthly and yearly data sets:

```{r}
head(cases_month)
head(cases_year)
```

## Data Cleaning

This data was cleaned using the **janitor** package to standardize messy Excel headers into snake_case.

For the **monthly** dataset, columns ranging from `year`to `discarded` were converted to numeric types to fix import errors where numbers were read as characters.

For the **annual** data set, there were header corrections, variable renaming, and row filtering. The first row of the spreadsheet was converted to become the column names using `row_to_names(1)`. Specific columns that imported with generic names (e.g., `na`, `na_2`) were manually mapped to their true categories, such as `measles_lab_confirmed` and `rubella_epi_linked`, to ensure the data is searchable and interpretable. The first row of data was removed using `slice(-1)` to delete the redundant first row.

## Explorations

**RQ 1:** Do some countries experience recurring outbreaks while others show sustained declines?

**RQ 2:** Do countries with lower vaccination rates experience significantly higher measles case counts compared to countries with high vaccination coverage?

**RQ 3:** Which countries experience disproportionately high measles case rates relative to their population size?

**Additional Data:** Population data by country (e.g., World Bank)

**RQ 4:** Do countries with higher healthcare spending or stronger immunization programs experience fewer measles outbreaks over time?

**Additional Data:** Healthcare expenditure or immunization coverage indicators (e.g., WHO / World Bank)

## Data Visualization 

Sketches - Two sketches of visuals *both* address RQs
