---
title: "Project 1"
author: "Nymph Kong"
format: html
editor: visual
---

```{r}
#| include: false
library(tidytuesdayR)
library(tidyverse)

tuesdata <- tidytuesdayR::tt_load('2025-06-24')
cases_month <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_month.csv')
cases_year <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-06-24/cases_year.csv')
summarize(cases_year, n_distinct(country))
```

Every time you make a change to a document (new or existing), you will need to follow these four steps:

1.  Stage the file—check the box next to the file.

2.  Write a commit message—describe what changes you made.

3.  Click “Commit” to log these changes.

4.  Click “Push” to send these changes up to GitHub.

## Data Context

This data contains measles and rubella cases across the world. This data was downloaded from the World Health Organisation [Provisional monthly measles and rubella data](https://immunizationdata.who.int/global?topic=Provisional-measles-and-rubella-data&location=) on 2025-06-12. Measles is a worldwide threat across many countries, with varying degrees of outbreaks. The data collects measles and rubella cases: clinically-compatible, epidemiologically linked, and laboratory-confirmed across 194 countries and separates the data into two data frames, monthly and yearly statistics. We can analysis this data to answer various important questions about theses diseases such as :

-   How have global measles cases changed over time?

-   Which regions or countries consistently report the highest measles burden?

-   Are there seasonal patterns in measles outbreaks across different regions?

-   Does the ratio of laboratory-confirmed cases to total cases reveal differences in healthcare capacity across countries?

## Data Cleaning

Data cleaning was done to better read the data and help when visualizing. Are datatypes changed? Are columns renamed? Are variables created or modified?

```{r}
# Data provided by WHO Provisional measles and rubella data. Cleaning variable
# names and fixing data types.

library(tidyverse)
library(here)
library(readxl)
library(janitor)

here::here()

cases_month <- read_xlsx(here("404-table-web-epi-curve-data.xlsx"), 2) %>%
  janitor::clean_names() %>%
  dplyr::mutate(
    dplyr::across(year:discarded, as.numeric)
  )

cases_year <- read_xlsx(here("403-table-web-reporting-data.xlsx"), 2) %>%
  row_to_names(1) %>%
  clean_names() %>%
  rename(
    country = member_state,
    iso3 = iso_country_code,
    measles_total= number_of_measles_cases_by_confirmation_method,
    measles_lab_confirmed = na,
    measles_epi_linked = na_2,
    measles_clinical = na_3,
    rubella_total = number_of_rubella_cases_by_confirmation_method,
    rubella_lab_confirmed = na_4,
    rubella_epi_linked = na_5,
    rubella_clinical = na_6
  ) %>%
  slice(-1)

```

Research Questions w/ Og Data 5 pts Poses Two Thoughtful RQ RQs aren't easily answered RQs include 2+ variables

Research Questions w/ Supplemental Data 5 pts Poses Two Thoughtful RQ RQs aren't easily answered RQs include 2+ variables

Data Visualization Sketches 5 pts Two sketches of visuals *both* address RQs
